---
title: "MCMC & confidence bands"
format: 
    revealjs:
      mainfont: Lato
      smaller: true
---

# Exercises

## Setup

Consider the following data generative model and priors:

$$
\begin{aligned}
Y_i | \beta_0, \beta_1, x_i &\sim N(\beta_0 + \beta_1x_i, \sigma^2)\\
\beta_0, \beta_1 &\sim \text{ iid } N(0, 2)\\
\sigma^2 &\sim \text{inverse-gamma}(2, 3)
\end{aligned}
$$

Tip: load `library(invgamma)` to use the `dinvgamma` function.

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(invgamma)
```

```{r}
#| echo: false
set.seed(602)
x = rnorm(30, 0, 10)
y = x * 2 + rnorm(30, 0, 3)

df = data.frame(y, x)
# write_csv(df, "simulatedXY.csv")
``` 


```{r}
#| eval: false
# load the data
YX = readr::read_csv("https://sta602-sp26.github.io/data/simulatedXY.csv")
y = YX$y
x = YX$x
```

## Code

Below is code to sample from

$$
p(\beta_1, \beta_2, \sigma^2 | y_1, \ldots, y_n, x_1, \ldots, x_n)
$$.

```{r}
set.seed(602)
logLikelihood = function(beta0, beta1, sigma2) {
  mu = beta0 + (beta1 * x)
  sum(dnorm(y, mu, sqrt(sigma2), log = TRUE))
}

logPrior = function(beta0, beta1, sigma2) {
  dnorm(beta0, 0, sqrt(2), log = TRUE) + 
    dnorm(beta1, 0, sqrt(2), log = TRUE) +
    dinvgamma(sigma2, 2, 3, log = TRUE)
}

logPosterior = function(beta0, beta1, sigma2) {
  logLikelihood(beta0, beta1, sigma2) + logPrior(beta0, beta1, sigma2)
}


BETA0 = NULL
BETA1 = NULL
SIGMA2 = NULL

accept1 = 0
accept2 = 0
accept3 = 0

S = 500

beta0_s = 0.1
beta1_s = 10
sigma2_s = 1
for (s in 1:S) {
  
  ## propose and update beta0
  beta0_proposal = rnorm(1, mean = beta0_s, .5)
   log.r = logPosterior(beta0_proposal, beta1_s, sigma2_s) - 
     logPosterior(beta0_s, beta1_s, sigma2_s)
   
   if(log(runif(1)) < log.r)  {
    beta0_s = beta0_proposal
    accept1 = accept1 + 1 
   }
   
   BETA0 = c(BETA0, beta0_s)
   
   ## propose and update beta1
    beta1_proposal = rnorm(1, mean = beta1_s, .5)
   log.r = logPosterior(beta0_s, beta1_proposal, sigma2_s) - 
     logPosterior(beta0_s, beta1_s, sigma2_s)
   
   if(log(runif(1)) < log.r)  {
    beta1_s = beta1_proposal
    accept2 = accept2 + 1 
   }
   
   BETA1 = c(BETA1, beta1_s)
   
   ## propose and update sigma2
   ### note: sigma2 is positive only, we want to only propose positive values
   sigma2_proposal = 1 / rgamma(1, shape = 1, sigma2_s)
   log.r = logPosterior(beta0_s, beta1_s, sigma2_proposal) - 
     logPosterior(beta0_s, beta1_s, sigma2_s) - 
     dinvgamma(sigma2_proposal, 1, sigma2_s, log = TRUE) +
     dinvgamma(sigma2_s, 1, sigma2_proposal, log = TRUE) 
   
   if(log(runif(1)) < log.r)  {
    sigma2_s = sigma2_proposal
    accept3 = accept3 + 1 
   }
   
   SIGMA2 = c(SIGMA2, sigma2_s)
   
}
```


## Exercise 1

- Is a Metropolis or Metropolis-Hastings algorithm used to sample $\sigma^2$? Hint: What is the proposal distribution used to sample $\sigma^2$? Is it symmetric? Can you show that it is or is not symmetric in R? 

- Create a trace plot for each parameter. How many iterations does it take for $\beta_1$ to converge to the posterior? What is going on with the $\sigma^2$ trace plot?

- Report the effective sample size of each parameter, $\beta_0$, $\beta_1$, and $\sigma^2$. Hint: load the `coda` package and check the lecture notes.

## Exercise 2 

- Go back and increase the length of the chain, $S$, until you obtain approximately 200 ESS for each parameter. Re-examine your plots and discuss.

- Report the posterior median and a 90% confidence interval for each unknown. Why is the median a better summary than the mean for $\sigma^2$?

## Exercise 3

- Plot the posterior mean $\theta(x)$ vs $x$ where $\theta(x) \equiv \beta_0 + \beta_1 x$.

- Additionally, plot a 95% confidence band for $\theta(x)$ vs $x$ and interpret your plot.

# Solutions

## Solution 1, 2

::: panel-tabset

## Algo.

A Metropolis-Hastings algorithm. inverse-gamma proposal is asymmetric. 

Showing with `R` code: 

```{r}
sigma2_proposal = 1.5
sigma2_s = 1
dinvgamma(sigma2_proposal, 1, sigma2_s) == 
  dinvgamma(sigma2_s, 1, sigma2_proposal)
```

## Trace plots (img)

```{r}
#| echo: false
library(patchwork)
parameterDF = data.frame(BETA0, BETA1, SIGMA2)
n = nrow(parameterDF)

p0 = parameterDF %>%
  ggplot(aes(x = 1:n)) +
  geom_line(aes(y = BETA0)) +
  theme_bw() +
  labs(x = "iteration", y = "beta0")

p1 = parameterDF %>%
  ggplot(aes(x = 1:n)) +
  geom_line(aes(y = BETA1)) +
  theme_bw() +
  labs(x = "iteration", y = "beta1")

p2 = parameterDF %>%
  ggplot(aes(x = 1:n)) +
  geom_line(aes(y = SIGMA2)) +
  theme_bw() +
  labs(x = "iteration", y = "sigma2")

p0 + p1 + p2
```

- $\beta_1$ converges to the target after about 50 iterations.

- Some samples from inv-gamma are huge, the huge variance results in a flatter likelihood and still get accepted.


## Trace plots (code)

```{r}
#| eval: false
library(patchwork)
parameterDF = data.frame(BETA0, BETA1, SIGMA2)
n = nrow(parameterDF)

p0 = parameterDF %>%
  ggplot(aes(x = 1:n)) +
  geom_line(aes(y = BETA0)) +
  theme_bw() +
  labs(x = "iteration", y = "beta0")

p1 = parameterDF %>%
  ggplot(aes(x = 1:n)) +
  geom_line(aes(y = BETA1)) +
  theme_bw() +
  labs(x = "iteration", y = "beta1")

p2 = parameterDF %>%
  ggplot(aes(x = 1:n)) +
  geom_line(aes(y = SIGMA2)) +
  theme_bw() +
  labs(x = "iteration", y = "sigma2")

p0 + p1 + p2
```

## ESS
```{r}
library(coda) 

parameterDF = data.frame(BETA0, BETA1, SIGMA2)
apply(parameterDF, 2, effectiveSize)

```

## Longer run

```{r}
set.seed(602)
logLikelihood = function(beta0, beta1, sigma2) {
  mu = beta0 + (beta1 * x)
  sum(dnorm(y, mu, sqrt(sigma2), log = TRUE))
}

logPrior = function(beta0, beta1, sigma2) {
  dnorm(beta0, 0, sqrt(2), log = TRUE) + 
    dnorm(beta1, 0, sqrt(2), log = TRUE) +
    dinvgamma(sigma2, 2, 3, log = TRUE)
}

logPosterior = function(beta0, beta1, sigma2) {
  logLikelihood(beta0, beta1, sigma2) + logPrior(beta0, beta1, sigma2)
}


BETA0 = NULL
BETA1 = NULL
SIGMA2 = NULL

accept1 = 0
accept2 = 0
accept3 = 0

S = 10000

beta0_s = 0.1
beta1_s = 10
sigma2_s = 1
for (s in 1:S) {
  
  ## propose and update beta0
  beta0_proposal = rnorm(1, mean = beta0_s, .5)
   log.r = logPosterior(beta0_proposal, beta1_s, sigma2_s) - 
     logPosterior(beta0_s, beta1_s, sigma2_s)
   
   if(log(runif(1)) < log.r)  {
    beta0_s = beta0_proposal
    accept1 = accept1 + 1 
   }
   
   BETA0 = c(BETA0, beta0_s)
   
   ## propose and update beta1
    beta1_proposal = rnorm(1, mean = beta1_s, .5)
   log.r = logPosterior(beta0_s, beta1_proposal, sigma2_s) - 
     logPosterior(beta0_s, beta1_s, sigma2_s)
   
   if(log(runif(1)) < log.r)  {
    beta1_s = beta1_proposal
    accept2 = accept2 + 1 
   }
   
   BETA1 = c(BETA1, beta1_s)
   
   ## propose and update sigma2
   ### note: sigma2 is positive only, we want to only propose positive values
   sigma2_proposal = 1 / rgamma(1, shape = 1, sigma2_s)
   log.r = logPosterior(beta0_s, beta1_s, sigma2_proposal) - 
     logPosterior(beta0_s, beta1_s, sigma2_s) - 
     dinvgamma(sigma2_proposal, 1, sigma2_s, log = TRUE) + 
     dinvgamma(sigma2_s, 1, sigma2_proposal, log = TRUE) 
   
   if(log(runif(1)) < log.r)  {
    sigma2_s = sigma2_proposal
    accept3 = accept3 + 1 
   }
   
   SIGMA2 = c(SIGMA2, sigma2_s)
   
}

parameterDF = data.frame(BETA0, BETA1, SIGMA2)
apply(parameterDF, 2, effectiveSize)
```

## Post medians

```{r}
#| echo: false
library(knitr)
pmeans = parameterDF %>%
  apply(2, median) %>%
  round(1)

lower = round(c(quantile(BETA0, 0.05)[[1]],
          quantile(BETA1, 0.025)[[1]],
          quantile(SIGMA2, 0.025)[[1]]), 1)

upper = round(c(quantile(BETA0, 0.95)[[1]],
          quantile(BETA1, 0.975)[[1]],
          quantile(SIGMA2, 0.975)[[1]]), 1)

tObj = rbind(pmeans, "CI" = paste0("(", lower, ", ", upper, ")")) %>%
  t()

rownames(tObj) = c("beta0", "beta1", "sigma2")
  
tObj %>%
  kable(col.names = c("post. median", "CI"))

```

:::


## Solution 3

::: panel-tabset

## plot
```{r}
#| echo: false
get_theta_CI = function(X) {
     f = BETA0 + (BETA1 * X)
     return(quantile(f, c(0.025, 0.975)))
}

get_theta_mean = function(X) {
  f = BETA0 + (BETA1 * X)
  return(mean(f))
}

xlo = min(x)
xhi = max(x)
xVal = seq(xlo, xhi, by = 0.01)
lower = NULL
upper = NULL
M = NULL
   
for (i in seq_along(xVal)) {
  theta_CI = get_theta_CI(xVal[i])
  lower = c(lower, theta_CI[[1]])
  upper = c(upper, theta_CI[[2]])
  M = c(M, get_theta_mean(xVal[i]))
}

df = data.frame(xVal, lower, upper, M)
df %>%
  ggplot(aes(x = xVal)) +
  geom_line(aes(y = lower)) +
  geom_line(aes(y = upper)) +
  geom_line(aes(y = M), col = "red") +
  theme_bw() +
  labs(y = "theta",
       x = "x")
```

The red line shows our posterior expectation of $\theta(x)$ for each $x$. 
The black bands show our 95% confidence interval $\theta(x)$.

## code

```{r}
#| echo: true
#| eval: false
get_theta_CI = function(X) {
     f = BETA0 + (BETA1 * X)
     return(quantile(f, c(0.025, 0.975)))
}

get_theta_mean = function(X) {
  f = BETA0 + (BETA1 * X)
  return(mean(f))
}

xlo = min(x)
xhi = max(x)
xVal = seq(xlo, xhi, by = 0.01)
lower = NULL
upper = NULL
M = NULL
   
for (i in seq_along(xVal)) {
  theta_CI = get_theta_CI(xVal[i])
  lower = c(lower, theta_CI[[1]])
  upper = c(upper, theta_CI[[2]])
  M = c(M, get_theta_mean(xVal[i]))
}

df = data.frame(xVal, lower, upper, M)
df %>%
  ggplot(aes(x = xVal)) +
  geom_line(aes(y = lower)) +
  geom_line(aes(y = upper)) +
  geom_line(aes(y = M), col = "red") +
  theme_bw() +
  labs(y = "theta",
       x = "x")
```
:::

<!-- ## Exercise 1 -->

<!-- Let $p(\theta_1, \theta_2 | \mathbf{y})$ be our **target distribution**, i.e. the distribution we are interested in sampling. -->

<!-- We construct a Gibbs sampler and look at the trace plots of $\theta_1$ and $\theta_2$, produced below. Chat with your neighbor, describe what you observe. Has the chain **converged** for each parameter? How well is the sampler **mixing**? Do you think the parameters are correlated or uncorrelated?  -->

<!-- Write down a description of the plots below, in 2 paragraphs or less, for a reader who has **not taken this class**. You may assume the reader has taken a course on probability (e.g. similar to STA230/240). -->


<!-- ```{r} -->
<!-- #| echo: false -->
<!-- #| warning: false -->
<!-- library(tidyverse) -->
<!-- library(latex2exp) -->
<!-- library(patchwork) -->
<!-- beta0 = c(rnorm(750, mean = 10, sd = 1), rnorm(250, mean = 50, sd = 4)) -->
<!-- beta1 = c(rnorm(750, mean = 45, sd = 2), rnorm(250, mean = 53, sd = 1)) -->

<!-- df = data.frame(beta0, beta1) -->
<!-- p1 = df %>% -->
<!-- ggplot(aes(x = 1:1000)) + -->
<!--   geom_line(aes(y = beta0)) + -->
<!--   theme_bw() + -->
<!--   labs(y = TeX("$\\theta_1$"), x = "iteration") -->

<!-- p2 = df %>% -->
<!-- ggplot(aes(x = 1:1000)) + -->
<!--   geom_line(aes(y = beta1)) + -->
<!--   theme_bw() + -->
<!--   labs(y = TeX("$\\theta_2$"), x = "iteration") -->

<!-- p1 + p2 -->
<!-- ``` -->

<!-- ## Exercise 2 -->

<!-- Based on the first 1000 iterations of your Gibbs sampler shown on the previous slide, which of the following joint densities is the most plausible for $\theta_1, \theta_2 | \mathbf{y}$? Why? Hint: it may help to think about where your sampler starts and imagine a particle moving through space according to conditional updates. -->

<!-- ```{r} -->
<!-- #| echo: false -->
<!-- #| warning: false -->

<!-- beta0_true = c(rnorm(3000, mean = 10, sd = 1), rnorm(1000, mean = 50, sd = 4)) -->
<!-- beta1_true = c(rnorm(3000, mean = 45, sd = 2), rnorm(1000, mean = 53, sd = 1)) -->

<!-- beta0_base =  c(rnorm(2000, mean = 10, sd = 1), rnorm(1000, mean = 50, sd = 4), -->
<!--                  rnorm(1000, mean = 80, sd = 3)) -->

<!-- beta1_base = c(rnorm(2000, mean = 45, sd = 2), rnorm(1000, mean = 53, sd = 1), -->
<!--                 rnorm(1000, mean = 35, sd = 2)) -->


<!-- df_base = data.frame(beta0 = beta0_base, beta1 = beta1_base) -->
<!-- df_true = data.frame(beta0 = beta0_true, beta1 = beta1_true) -->

<!-- p1 = df_true %>% -->
<!--   mutate(beta0 = beta0 / 10, -->
<!--          beta1 = beta1 / 10) %>% -->
<!--   ggplot(aes(x = beta0, y = beta1)) +  -->
<!--   geom_bin2d(bins = 25) + -->
<!--   theme_bw() +  -->
<!--   labs(y = TeX("$\\theta_2$"),  -->
<!--        x = TeX("$\\theta_1$")) -->

<!-- df2 = df_base -->

<!-- df2$beta0 = rnorm(4000, mean = 50, sd = 4) -->

<!-- p2 = df2 %>% -->
<!--   ggplot(aes(x = beta0, y = beta1)) + -->
<!--   geom_bin2d(bins = 25) + -->
<!--   theme_bw() + -->
<!--   labs(y = TeX("$\\theta_2$"), -->
<!--        x = TeX("$\\theta_1$")) -->

<!-- df3 = df_base -->
<!-- df3$beta0 = rnorm(4000, mean = 35, sd = 15) -->
<!-- df3$beta1 = rnorm(4000, mean = 48, sd = 10) -->


<!-- p3 = df3 %>% -->
<!--   ggplot(aes(x = beta0, y = beta1)) + -->
<!--   geom_bin2d(bins = 25) + -->
<!--   theme_bw() + -->
<!--   labs(y = TeX("$\\theta_2$"), -->
<!--        x = TeX("$\\theta_1$")) -->

<!-- df4 = df_base -->

<!-- p4 = df4 %>% -->
<!--   ggplot(aes(x = beta0, y = beta1)) + -->
<!--   geom_bin2d(bins = 25) + -->
<!--   theme_bw() + -->
<!--   labs(y = TeX("$\\theta_2$"), -->
<!--        x = TeX("$\\theta_1$")) -->


<!-- (p1 + p2) / (p3 + p4) + -->
<!--   plot_annotation(title = "Joint densities", tag_levels = c("A")) -->
<!-- ``` -->

<!-- ## Exercise 3 -->



