---
title: "Homework 4"
subtitle: "Due Friday February 13 at 5:00pm"
mainfont: Lato
format: 
  html:
    toc: true
---


## Exercise 1

Let $\theta$ be the rate of mutation for a certain cluster of cancer cells. Biologists encode their prior uncertainty about $\theta$ with the following density:

$$
p(\theta) = \frac{4}{10}e^{-4\theta} + \frac{9}{10 \Gamma(8)} \theta^7 e^{-\theta}
$$
a. Make a plot of this prior density and explain what it means, in context, to the biologists.

b. Let $Y_i$ be the number of mutations produced in the $i$th cell, such that

$$
Y_i | \theta \sim \text{iid }Poisson(\theta)
$$

Write out the posterior distribution of $\theta$ given $y_1,\ldots y_n$ (up to a proportionality constant) and simplify as much as possible. Hint: Be careful when writing your proportionality statement!

c. The posterior is a mixture (weighted average) of two distributions that you know. Identify these two distributions, including their parameters.

d. Assume you obtain mutation data from two cells, $y_1 = 3, y_2 = 1$. Compute the posterior exactly (i.e. find the appropriate integration constant) and plot the posterior density. 

e. In part (c) you identified the posterior as a mixture (weighted average) of two distributions. Given the data in part (d), compute the weights of each density in the mixture.

## Exercise 2

A group of scientists have mutation data, 

```{r}
y = c(0,3,0,1,5,2,0,4,1,1)
```

and are interested in assessing how well the Poisson model from exercise 1 fits their data. Using the data generative model and prior from exercise 1, generate posterior predictive datasets $y^{(1)}, y^{(2)}, \ldots y^{(S)}$, where each data set $y^{(s)}$ is a vector of length 10 whose entries are sampled from the Poisson distribution with parameters $\theta^{(s)}$. Each $\theta^{(s)}$ itself is a sample from the posterior $p(\theta | y_1, \ldots y_{10})$. For each $s$, let $t(s)$ be the sample average of the 10 values of $y^{(s)}$ divided by the sample standard deviation of $y^{(s)}$. Make a histogram of $t(s)$ and compare to the observed value of this statistic. Based on this statistic, assess the fit of the Poisson model for these data.


## Exercise 3

Chimowitz et al. (2011) [https://doi.org/10.1056/NEJMoa1105335](https://doi.org/10.1056/NEJMoa1105335) investigate if stents are effective treatment to manage strokes in patients with atherosclerotic intracranial arterial stenosis. You can load the data using the code below. Data sourced from `openintro` package.

```{r}
#| eval: false
#| message: false
stents = readr::read_csv("https://sta602-sp26.github.io/data/stent365.csv")
```

Each row of the data set is an individual patient. The `group` column indicates whether the patient was treated with a stent or not. The `outcome` column reports whether the patient had a stroke or not within a year.

a. Write down a data generative model for this data. Hint: you might write down two data generative models, one for patients in the treatment group and one for patients in the control group.

b. Write down your prior beliefs about unknown parameters in your model above using a conjugate prior. Choose parameters for the priors. Explain your choices.

c. Report (using Monte Carlo sampling or otherwise) the posterior mean of the [relative risk](https://en.wikipedia.org/wiki/Relative_risk), i.e. the posterior mean of the probability of stroke in the treatment group versus in the control, (think $E~[\frac{\theta_t}{\theta_c}~|~\text{data}]$). Additionally, include a 95% posterior confidence interval for the relative risk.

d. Plot the posterior of the relative risk from part c. Do you believe the treatment is effective?

## Exercise 4 

Let 

$$
\begin{aligned}
Y | \theta, \gamma &\sim N(\theta, 1/\gamma)\\
\theta | \gamma &\sim N(\mu_0, 1/ (\gamma \kappa_0))\\
\gamma &\sim \text{gamma}(a, b)
\end{aligned}
$$
so $\gamma$ is the precision (inverse-variance) of the normal distribution.

a. Derive and simplify the joint pdf $p(y_1, \ldots y_n | \theta, \gamma)$.

b. Derive the marginal posterior of the precision, $p(\gamma| y_1, \ldots y_n)$.

c. Explain the drawback(s), if any, to specifying $p(\theta, \gamma) = p(\theta) p(\gamma)$, where $\theta \sim N(\mu_0, \tau_0)$ and $\gamma \sim \text{gamma}(a,b)$.

<!-- *Bayes estimation under quadratic loss* -->

<!-- Let $Y \mid \theta \sim \mathcal{N}(\theta, \sigma^2)$ with known $\sigma^2$, and suppose the prior distribution is -->
<!-- $$ -->
<!-- \theta \sim \mathcal{N}(\mu_0, \tau^2). -->
<!-- $$ -->

<!-- a. Write down the distribution of $\theta \mid Y = y$. You can write "the posterior is distributed BLANK with parameters BLANK and BLANK, (don't need to derive, can follow class notes). -->

<!-- b. Fix $\mu_0 = 0$, $\tau^2 = 1$, $\sigma^2 = 1$, and $y = 1$. Simulate a large sample $\{\theta^{(1)}, \dots, \theta^{(S)}\}$ from the posterior distribution $p(\theta \mid y)$. -->

<!-- c. Consider estimators of $\theta$ of the form $\delta_c = c$, where $c \in \mathbb{R}$. Using your simulated posterior draws, approximate the posterior expected quadratic loss -->
<!--    $$ -->
<!--    R(c) = \mathbb{E}\!\left[(\theta - c)^2 \mid Y = y\right] -->
<!--    $$ -->
<!--    over a grid of values of $c$. -->

<!-- d. Plot the estimated risk $R(c)$ as a function of $c$.   -->
<!--    Identify the value of $c$ that minimizes the empirical risk, and compare it to the posterior mean derived in part (a).   -->



```{r}
#| echo: false
#| eval: false
## Empirical Bayes estimator under quadratic loss via posterior simulation

set.seed(1)

# Given values
mu0   <- 0
tau2  <- 1
sig2  <- 1
y     <- 1
S     <- 10000

# Posterior for Normal-Normal model:
# theta|y ~ N(m_post, v_post)
v_post <- 1 / (1/tau2 + 1/sig2)
m_post <- v_post * (mu0/tau2 + y/sig2)

# Simulate posterior draws
theta <- rnorm(M, mean = m_post, sd = sqrt(v_post))

# Empirical posterior risk for constant decision c under quadratic loss
c_grid <- seq(m_post - 3*sqrt(v_post), m_post + 3*sqrt(v_post), length.out = 500)
risk   <- sapply(c_grid, function(c) mean((theta - c)^2))

# Minimizer (empirical)
c_hat <- c_grid[which.min(risk)]

# Plot risk curve and compare to posterior mean
plot(c_grid, risk, type = "l", xlab = "c", ylab = "Estimated E[(theta - c)^2 | y]")
abline(v = m_post, lty = 2)     # posterior mean (theory)
abline(v = c_hat,  lty = 3)     # empirical minimizer

cat("Posterior mean (theory):", m_post, "\n")
cat("Empirical minimizer c*:  ", c_hat, "\n")

```

