---
title: "Homework 7"
subtitle: "Due Friday March 27 at 5:00pm"
mainfont: Lato
format: 
  html:
    toc: true
---

## Exercise 1

Ridge regression theory: Let $y \sim N_n(X \beta, \sigma^2 I)$. Consider estimating $\beta$ with the prior distribution $\beta | \sigma^2 \sim N_p(0, \sigma^2 I / \lambda)$, where $\lambda$ is known and $\beta$ and $\sigma^2$ are unknown.

a. Derive the conditional distribution of $\beta | y, \sigma^2$ and, in particular, show that $E[\beta | y] = (X^TX + I \lambda)^{-1} X^Ty$. Denote this expectation $\hat{\beta}_\lambda$, which we can use as an estimator of $\beta$. What happens to $\hat{\beta}_\lambda$ as $\lambda \rightarrow 0$?

b. Consider the special case that $X^TX$ is a diagonal matrix with entries $x_1^T x_1, \ldots, x_p^T x_p$. Write down the mathematical expression for an individual entry of the estimator, i.e. $\hat{\beta}_{\lambda~i}$ (where $i$ is the $i$th entry of the vector). 
Further write down the mathematical expression for individual entries of the OLS estimator $\hat{\beta}_{OLS~i}$.
Compare the two and explain, in words, the effect of $\lambda$. 

## Exercise 2

Ridge regression application: The data set `yX.diabetes.train` contains data on diabetes progression (first column) and 64 predictor variables. These data can be loaded with with command

```{r}
#| eval: false
yX<-dget(url("https://sta602-sp26.github.io/data/yX.diabetes.train"))
```

a. For each value of $\lambda \in \{0, 1,  \ldots, 99, 100 \}$ compute the estimator $\hat{\beta}_{\lambda}$ based on exercise 1 above. Visualize each $\hat{\beta}$ as a function of $\lambda$ (maybe using `matplot`). Describe any trend(s) you notice in the plots.

b. Load the data set `yX.diabetes.test` using the code below

```{r}
#| eval: false
yX.diabetes.test<-dget(
  url("https://sta602-sp26.github.io/data/yX.diabetes.test"))
```

Use `yX.diabetes.test` to evaluate the predictive performance of each estimate you obtained in part a. Specifically, compute the predictive error sum of squares $PSS(\lambda) = ||y_{test} - X_{test} \hat{\beta}_{\lambda}||^2$ for each value of $\lambda$ (IMPORTANT: $\hat{\beta}_{\lambda}$ is obtained from the training data in part a, not the test data). Make a plot of PSS versus $\lambda$. How good is the unbiased OLS estimate for prediction, relative to the other estimates?

c. Identify the value of $\lambda$ that has the best predictive performance. For this best value of $\lambda$, report which x-variables have the largest effects.

## Exercise 3

For this exercise, use the code below to load the data

```{r}
#| eval: false
yX = readRDS(
  url("https://sta602-sp26.github.io/data/yXspectroscopy.rds"))
```


Source separation: The first column y of the dataset `yXSS.rds` is the vectorization of a spectroscopy image of a water sample taken from the Neuse River in North Carolina. You can view the image with the following code: `y<-yX[,1] ; image(matrix(y,151,43))`. The water sample is of unknown origin, but it is assumed that it is a mix of water from 9 different categories, whose average spectroscopy images are given by the remaining 9 columns $X$ of `yX`. You can view these images with the same code above, applied to each column of $X$.

a. From $y$ and $X$, infer the sources of the water sample using the linear model $E[y|X, \beta] = X\beta$. Assuming the normal linear model and with priors $\beta \sim N_9(1/9, \ldots 1/9), I_9)$, $1/\sigma^2 \sim \text{gamma}(1, 1)$, use a Gibbs sampler to obtain a posterior distribution of $\beta$ and $\sigma^2$ given $y$. Plot the posterior density of $\sigma^2$, and obtain posterior 95% confidence intervals for each element of $\beta$. Which of the nine categories are the main sources of the water sample?

c. Evaluate the assumptions of the normal linear model using some residual plots, addressing the assumption that the entries of $y$ have constant variance, are uncorrelated, and are normally distributed. Hint: plots residuals vs yhat and a QQ plot.

d. For this problem it doesn't make sense for the coefficients of $\beta$ to be negative. Think of a modification to the prior distribution for $\beta$ that takes this fact into account, and describe *how you could* sample $\beta$ under this updated model (you don't have to implement it).