---
title: "Bayesian model averaging"
author: "Dr. Alexander Fisher"
# mainfont: Lato
format: 
  html:
    toc: true
execute:
  warning: false
  message: false
---

## Motivation

Consider the standard regression problem,

$$
Y | \beta, \sigma^2 \sim N_n(X\beta, \sigma^2 I_n)
$$
where each term is of typical dimension,

- $Y \in \mathbb{R}^n$ vector of observations, 
- $X \in \mathbb{R}^{n \times p}$ matrix of predictors, 
- $\beta \in \mathbb{R}^p$ vector of coefficients,
- $\sigma^2 \in \mathbb{R}$, and
- $I_n \in \mathbb{R}^{n \times n}$.

Often, when we have many covariates, i.e. $p$ is large, we believe that some do not really explain $Y$. A particular subset of the explanatory variables is termed a **model** of $Y$. Since each explanatory variable belongs in the model, or does not, there are $K = 2^p$ total models we can consider. 

For notational convenience, let $M_1, \ldots, M_{K}$ be the models under study.

:::callout-important
For even modestly large $p$, such as $p=64$ in the example below, there are more possible models $2^{64} \approx 1.8 \times 10^{19}$ than there are grains of sand on earth (about $7.5 \times 10^{18}$). 

For this reason, we will never be able to sample a substantial proportion of all the models.
:::

There are many different questions one might answer with a regression model. Sometimes, the model itself is of interest i.e. we wish to select a single model $M_j$. Other times, we may care about estimating a particular coefficient $\beta_i$ to understand how a single predictor relates to $Y$. Alternatively, we may really care about making predictions for new data.

These tasks are not mutually exclusive, but might direct our approach differently.

In particular, if we are interested in prediction, or estimating a particular coefficient, then it's important to realize that estimates such as $\hat{\beta}$ or a function thereof, $\hat{Y} = X \hat{\beta}$ are implicitly conditioned on a specific model.

- How can we address uncertainty in the model itself?

## Model averaging setup

Let $z \in \mathbb{R}^p$ be a vector of indicator variables $z_1, \ldots, z_p$, where $z_j = 1$ means the $j$th predictor is in the model, and $z_j = 0$ means it is not. Conceptually, $z$ tells us which regression coefficients are non-zero.

$$
y_i = z_1 \beta_1 x_{i, 1} + \ldots + z_p \beta_p x_{i,p} + \epsilon
$$

where $\epsilon \sim N(0, \sigma^2)$.

In matrix form,

$$
Y|\beta, z, \sigma^2 \sim N_n(XZ\beta, \sigma^2 I_n)
$$

where $Z$ is a diagonal matrix with components $Z_{ii} = z_i$.

::: callout-note
Each separate "model" $M$ is equivalent to a different vector $z$.
:::

Once we've setup our likelihood to explicitly condition on the model, $z$, we continue our usual Bayesian practice: we make inferences about *all* unknowns: $\beta, \sigma^2$, and $z$.

Now, however we can differentiate between estimates of $\beta$ conditional on a model,

$$
E[\beta | y, z] = \int \beta p(\beta | y, z) d\beta
$$ 

versus estimates of $\beta$ that account for model uncertainty: 

$$
\begin{aligned}
E[\beta | y] = \sum_{z} \int \beta p(\beta, z | y) d\beta
\end{aligned}
$$

## Complete Bayesian model 

::: panel-tabset

## Model

Data generative model: 

$$
Y|\beta, z, \sigma^2 \sim N_n(X_z\beta_z, \sigma^2 I_n)
$$
Common prior specification:

$$
\begin{aligned}
\beta_z | \sigma^2 &\sim N_z \left(0, g \sigma^2 (X_z^TX_z)^{-1}\right)\\
1/\sigma^2 &\sim \text{gamma}(\frac{\nu_0}{2}, \frac{\nu_0}{2} \sigma_0^2)\\
p(z_i = 1) &= \frac{1}{2} \text{ for all i}
\end{aligned}
$$
where $g$ is some fixed constant and $\beta_z$ is the set of $\beta$ corresponding to non-zero entries of $z$ ($X_z$ is constructed similarly). The prior on $\beta$ is known as Zellner's $g$ prior [^1]. The $g$ prior is mathematically convenient and leads to nice marginal distributions for our Gibbs sampler.

Where does the prior variance of $\beta$ come from? Recall: $Var[\hat{\beta}_{OLS}] = \sigma^2 (X^TX)^{-1}$. So a priori we expect the variance of $\beta$ to look like the OLS variance and $g$ controls the weight of this prior variance (or $1/g$ controls the weight of the precision). Small $g$ implies more prior precision relative to the data. 

- Question: what should $g$ be equal to for unit precision?


<!-- The precision then, for $n$ observations is $(X^TX) / \sigma^2$. Unit precision would be $\frac{1}{n} \cdot (X^TX) / \sigma^2$.  -->

[^1]: Zellner, Arnold. "On assessing prior distributions and Bayesian regression analysis with g-prior distributions." Bayesian inference and decision techniques (1986).

## Gibbs updates

### Beta

$$
\beta_z | X_z, \sigma^2, y \sim N_z\left( A^{-1}b,
A^{-1}
\right)
$$
where $A^{-1} = \sigma^2 \left( \frac{g}{g+1}\right)(X_z^TX_z)^{-1}$ and $b = \frac{1}{\sigma^2} X_z^Ty$.

We can simplify the full conditional notation,

$$
\beta_z | X_z, \sigma^2, y \sim N_z\left(
\left(\frac{g}{g+1}\right)\hat{\beta}_{z(OLS)},~ 
\left(\frac{g}{g+1}\right)  \sigma^2 \left[X_z^TX_z\right]^{-1}
\right)
$$

### sigma-squared

We can integrate out $\beta$ analytically, for a more efficient sampler,

$$
1/\sigma^2 | X_z, y \sim \text{gamma}([\nu_0 + n]/2, [\nu_0 \sigma_0^2 + SSR_g^z]/2)
$$
where $SSR_g^z = y^T(I - \frac{g}{g+1} X_z (X_z^T X_z)^{-1} X_z^T)y$

### z

$$
z_j \sim \text{binary}(o_j/(1 + o_j))
$$
where $p_j$ is the probability that $z_j = 1$, and is given by 

$$
o_j = \frac
{Pr(z_j = 1 | y, X, z_{-j})}
{Pr(z_j = 0 | y, X, z_{-j})}
= \frac{p(y|X, z_{-j}, z_j = 1) p(z_j = 1)}
{p(y|X, z_{-j}, z_j = 0) p(z_j = 0)}
$$
See page 165 of the book for details on how to compute the marginal likelihood $p(y|X, z)$. 
:::

## Example: Diabetes

This example comes from Hoff Ch 9.

The data below contains our outcome variable, diabetes progression (first column) and 64 predictors. This example and data set appeared on a previous homework.

```{r}
#| code-fold: true
#| code-summary: load libraries
library(tidyverse)
library(mvtnorm)
library(latex2exp)
library(coda)
```


```{r}
#| code-fold: true
#| code-summary: data load and setup
yX.diabetes.train<-dget(url("https://sta602-sp26.github.io/data/yX.diabetes.train"))

X = yX.diabetes.train[,-1]
y = yX.diabetes.train[,1]
```

```{r}
#| code-fold: true
#| code-summary: useful functions

# log-probability of y given x
lpyx = function(y, X, g=length(y), 
                nu0=1, s20 = try(summary(lm(y~1+X))$sigma^2, silent = TRUE)) {
  n=nrow(X)
  p=ncol(X)
  if(p==0) {
    Hg = 0
    s20 =mean(y^2)
  }
  if(p>0) {
    Hg = (g / (g + 1)) * X %*% solve(t(X) %*% X) %*% t(X)
  }
  
  SSRg = t(y) %*% (diag(1, nrow = n) - Hg) %*%y
  -.5 *(n *log(pi) + p * log(1 + g) + (nu0 + n) * log(nu0 *s20 + SSRg)- nu0 *log(nu0 *s20)) +
    lgamma((nu0 + n) / 2) - lgamma(nu0 / 2
    )
}

SSR = function(y, X, g = length(y)) {
  n=nrow(X)
  p=ncol(X)
  Hg = (g/(g+1)) * X %*% solve(t(X) %*% X) %*% t(X) 
  return(t(y) %*% (diag(1,nrow=n) - Hg) %*% y)
}

getVarianceBeta = function(X, s2, g = length(y)) {
  (g / (g+1)) * s2 * solve(t(X) %*% X)
}

getOLS_mean = function(y, X, varBeta, g = length(y)) {
  varBeta %*%  t(X) %*% y
}

```


```{r}
#| code-fold: true
#| code-summary: MCMC
#| eval: false
# setup
S = 200 # num iterations

# fixed quantities
nu0 = 1
sigma02 = 1
alpha_n = (nu0 + n)/2
beta_const = nu0 * sigma02
g = length(y)
# objects to fill
Z = matrix(NA, S, p)
BETA = matrix(NA, S, p)
SIGMA2 = matrix(NA, S, 1)

## starting values ##
z = rep(1, ncol(X))
lpy.c = lpyx(y , X[,z==1, drop=FALSE])
gamma = 1
## Gibbs sampler
for(s in 1:S) {
  # update z
  for (j in sample(1:ncol(X))) {
    zp = z
    zp[j] = 1 - zp[j]
    lpy.p = lpyx(y, X[,zp==1, drop = FALSE])
    r = (lpy.p - lpy.c) * (-1)^(zp[j]==0)
    z[j] = rbinom(1, 1, 1 / (1+exp(-r)))
    if(z[j] == zp[j]) {
      lpy.c = lpy.p
    }
  }
  Z[s,] = z
  
  ## update sigma^2
  s2 = 1 / rgamma(1, alpha_n, (beta_const + 
                                 SSR(y, X[,z==1, drop = FALSE]))/2)
  
  SIGMA2[s] = s2
  
  ## update beta
  
  varBeta = getVarianceBeta(X[,z==1, drop = FALSE], s2)
  meanBeta = getOLS_mean(y, X[,z==1, drop = FALSE], varBeta / s2)
  
  beta = rmvnorm(1, mean = meanBeta,
                 sigma = varBeta)
  BETA[s,z==1] = beta
}


```

```{r}
#| echo: false
# write.table(Z, file="../data/Z-model-avg.txt", row.names=FALSE, col.names=FALSE)
# write.table(BETA, file="../data/BETA-model-avg.txt", row.names=FALSE, col.names=FALSE)
# write.table(SIGMA2, file="../data/SIGMA2-model-avg.txt", row.names=FALSE, col.names=FALSE)

Z = read.table("../data/Z-model-avg.txt")
BETA = read.table("../data/BETA-model-avg.txt")
SIGMA2 = read.table("../data/SIGMA2-model-avg.txt")
S = nrow(SIGMA2)
```


::: panel-tabset
## Probability of regressors

```{r}
#| echo: false
Z_df = data.frame(zval = Z %>%
                    colSums() / S)
# colnames(Z_df) = paste0("z", 1:64)
Z_df %>% 
  mutate(z = 1:64) %>%
  ggplot(aes(x = z, y = zval)) +
  geom_col() +
  theme_bw() +
  labs(y = TeX("$p(z_j = 1| y, X)$"), x = "regressor index")
```

## Diagnostics

### trace plots

```{r}
#| code-fold: true
data.frame(sigma2 = SIGMA2[,1]) %>%
  ggplot(aes(x = 1:S)) +
  geom_line(aes(y = sigma2)) +
  theme_bw() +
  labs(x = "iteration", y = TeX(r'($\sigma^2$)'))
```

```{r}
#| code-fold: true
data.frame(beta2 = BETA[,2]) %>%
  mutate_at(1, ~ replace(., is.na(.), 0)) %>%
  ggplot(aes(x = 1:S)) +
  geom_line(aes(y = beta2)) +
  theme_bw() +
  labs(x = "iteration", y = TeX(r'($\beta_2$)'))
```



### effective size
```{r}
apply(Z, 2, effectiveSize)
apply(SIGMA2, 2, effectiveSize)
```

## Posterior summaries

```{r}
#| echo: false
#| eval: false
# notes
postMeanZ = as.vector(colSums(Z) / S)
postMeanBeta = as.vector(colSums(BETA, na.rm = TRUE)) / as.vector(colSums(Z))
betaHat2 = postMeanBeta * postMeanZ
betaHat = as.vector(colSums(BETA, na.rm = TRUE) / S)
betaHat == betaHat2
```

```{r}
#| code-fold: true
yX.diabetes.test<-dget(
  url("https://sta602-sp26.github.io/data/yX.diabetes.test"))
Xtest = yX.diabetes.test[,-1]
ytest = yX.diabetes.test[,1]

betaHat = as.vector(colSums(BETA, na.rm = TRUE) / S)
yHat = Xtest %*% betaHat

data.frame(yHat, ytest) %>%
  ggplot(aes(x = yHat, y = ytest)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  theme_bw() +
  labs(x = TeX(r'($\hat{y}$)'), y = "y")
```

The predictive sum of squared error is 

```{r}
sum((yHat - ytest)^2)
```

Compare this to the predictive sum of squared error from ridge regression on the homework.
:::

