---
title: "Prediction checks and Monte Carlo Error"
author: "Dr. Alexander Fisher"
# mainfont: Lato
format: 
  html:
    toc: true
---

```{r}
#| warning: false
#| message: false
# load packages
library(tidyverse)
library(latex2exp)
```

## Predictive distributions with Monte Carlo

**Goal**: evaluate our model both *a priori* and *a posteriori* i.e. before and after looking at the data.

### Prior predictive distribution

We can use Monte Carlo to sample new observation, $\tilde{y}$, from the **prior predictive distribution**

$$
p(\tilde{y}) = \int p(\tilde{y}|\theta)p(\theta) d\theta,
$$


where we proceed by following the iterative procedure below

```
1. sample theta_i from the prior p(theta)
2. sample ytilde from p(ytilde | theta_i)
3. repeat steps 1 and 2
```

- this can be useful to see if a prior for $p(\theta)$ actually translate to reasonable prior beliefs about the data.

::: panel-tabset
## Exercise
For $p(\theta) = \text{gamma}(8,2)$, plot $p(\tilde{y})$ assuming $\tilde{y} | \theta \sim \text{Poisson}(\theta)$.
:::

```{r}
#| eval: false
#| echo: false

N = 10000

theta.mc = rgamma(N, 8, 2)
ytilde.mc = rpois(N, theta.mc)
hist(ytilde.mc)
```


### Posterior predictive distribution

We can also sample $\tilde{y}$ from the **posterior predictive distribution**,

$$
p(\tilde{y} | y_1, \ldots y_n) = \int p(\tilde{y}|\theta) p(\theta|y_1, \ldots, y_n)d\theta,
$$

where the procedure is the same as before, except step 1 is replace with sampling $\theta$ from the posterior $p(\theta | y_1,\ldots, y_n)$.

The resulting sequence $\{(\theta^{(1)}, \tilde{y}^{(1)}), \ldots, (\theta^{(S)}, \tilde{y}^{(S)})\}$ constitutes $S$ independent samples from the joint posterior of $(\theta, \tilde{Y})$. The sequence $\{\tilde{y}^{(1)}, \ldots, \tilde{y}^{(S)}\}$ constitutes $S$ independent samples from the marginal posterior distribution of $\tilde{Y}$, aka the posterior predictive distribution.

## Posterior predictive model checking

We can assess the fit of a model by comparing the posterior predictive distribution to the empirical distribution.

### Example: is our Poisson model flawed?

```{r}
#| warning: false
# load general social survey data
gss = read_csv("https://sta602-sp26.github.io/data/gss.csv")
```


```{r}
y1 = gss$CHILDS[gss$FEMALE == 1 &  gss$YEAR >= 1990  & gss$AGE == 40 & 
                   gss$DEGREE < 3 ]
y1 = y1[!is.na(y1)]
n = length(y1)
```

We are examining the number of children $Y_i$ belonging to $n=$ `r n` 40 year old women surveyed 1990 or later without a bachelor's. These data come from the general social survey.

Suppose

$$
\begin{aligned}
Y_i | \theta & \sim \text{Poisson}(\theta)\\
\theta & \sim \text{gamma}(2, 1).
\end{aligned}
$$
The empirical and predictive distributions of the data are both plotted below.

::: panel-tabset

## plot

```{r}
#| echo: false
#| warning: false

set.seed(123)

# posterior predictive distribution
ytotal = sum(y1)
a = 2 ; b = 1
N = 10000
theta.post.mc = rgamma(N, ytotal + a, b + n)
y1.mc = rpois(N, theta.post.mc)

# data
df = data.frame(y1) # empirical
df2 = data.frame(y1.mc) # post predictive
  
# make plot
df %>%
  ggplot(aes(x = y1)) +
  geom_bar(aes(x = y1 + .15, y = (..count..)/sum(..count..),
               fill = "empirical"), alpha = 0.6, width = 0.3) +
  geom_bar(data = df2, 
                 aes(x = y1.mc -.15, y = (..count..) / sum(..count..),
                     fill = "predictive"), alpha = 0.6, width = 0.3) +
  labs(x = "number of children", 
       y = TeX("$p(Y_i = y_i)$"),
       fill = "") +
  scale_x_continuous(breaks = c(0:7), labels = c(0:7),
                     limits = c(-.5,7.5)) +
  
  theme_bw()
```

## code

```{r}
#| eval: false
set.seed(123)

# posterior predictive distribution
ytotal = sum(y1)
a = 2 ; b = 1
N = 10000
theta.post.mc = rgamma(N, ytotal + a, b + n)
y1.mc = rpois(N, theta.post.mc)

# data
df = data.frame(y1) # empirical
df2 = data.frame(y1.mc) # post predictive
  
# make plot
df %>%
  ggplot(aes(x = y1)) +
  geom_bar(aes(x = y1 + .15, y = (..count..)/sum(..count..),
               fill = "empirical"), alpha = 0.6, width = 0.3) +
  geom_bar(data = df2, 
                 aes(x = y1.mc -.15, y = (..count..) / sum(..count..),
                     fill = "predictive"), alpha = 0.4, width = 0.3) +
  labs(x = "number of children", 
       y = TeX("$p(Y_i = y_i)$"),
       fill = "") +
  scale_x_continuous(breaks = c(0:7), labels = c(0:7),
                     limits = c(-.5,7.5)) +
  
  theme_bw()
```

:::


::: panel-tabset 

## Exercise
Let $\mathbf{y}$ be a vector of length 111. Let $t(\mathbf{y})$ be the ratio of $2$s to $1$s in $\mathbf{y}$. For our observed data, this test statistic $t(\mathbf{y}_{obs}) = 38 / 19 = 2$. What is the tail probability $p(t(\tilde{\mathbf{Y}}) \geq t(\mathbf{y}_{obs}))$ under the posterior predictive distribution?

## Solution
```{r}
set.seed(123)

t.mc = NULL
for (i in 1:10000) {
  theta1 = rgamma(1, ytotal + a, b + n) # draw 1 theta from posterior
  y1.mc = rpois(n, theta1) # draw y from post pred of n = 111
  t.mc = c(t.mc,
           sum(y1.mc == 2) / sum(y1.mc == 1))# compute t
}
hist(t.mc)
mean(t.mc >= 2)
```

:::

## Monte Carlo error


#### How many values should we simulate?

Recall: expected values are integrals, and integrals are expected values. Since central limit theorem (CLT) deals with expected values...

Recall: CLT states that if $\theta_i |\vec{y}$ iid with mean $\theta$ and finite variance $\sigma^2$, for $i \in \{1, \ldots, N\}$, then the sample mean

$$
\bar{\theta} \sim N(\theta, \frac{\sigma^2}{N} ).
$$

- How to remember this/show this? Offline notes.

So to estimate $\theta$, we can generate $\bar{\theta}$ by Monte Carlo simulation and report a confidence interval using quantiles of the normal given above in conjunction with the Monte Carlo standard error $\frac{\hat{\sigma}}{\sqrt{N}}$

This means we get convergence at the rate $\mathcal{O}\left(\frac{1}{\sqrt{N}}\right)$ regardless of the dimension of the integral!

Recall: 

```{r}
sd1 = pnorm(1) - pnorm(-1)
sd2 = pnorm(2) - pnorm(-2)
sd3 = pnorm(3) - pnorm(-3)
```

- a `r sd1`% confidence interval can be obtained using $\pm 1\cdot \hat{\sigma}/\sqrt{N}$
- a `r sd2`% confidence interval can be obtained using $\pm 2\cdot \hat{\sigma}/\sqrt{N}$
- a `r sd3`% confidence interval can be obtained using $\pm 3\cdot \hat{\sigma}/\sqrt{N}$

### Toy example

Let $\phi$ be the random variable of interest. 

Assume the posterior distribution $\phi| data \sim \text{binomial}(20, 0.4)$

```{r}
set.seed(123)

# binomial(n, p)
n = 20
p = 0.4

# exact posterior mean, variance, sd of a binomial(n, p)
EPhi = n*p # 20*.4 = 8
VarPhi = n*p*(1-p) # 20*.4*.6 = 4.8
sdPhi = sqrt(VarPhi) # 2.19089

# Monte Carlo sample of size N
N = 100
PhiSamples = rbinom(N, size = n, prob = p) 

# Empirical approximation of posterior mean, var, sd
PhiBar = mean(PhiSamples)
PhiVar = var(PhiSamples)
PhiSigma = sd(PhiSamples) # = sqrt(sum((xSamples - xbar)^2) / (N -1))

se = PhiSigma / sqrt(N)

lb = round(PhiBar - (2*se), 3)
ub = round(PhiBar + (2*se), 3)
```

For `r paste("N = ", N)` Monte Carlo samples, The posterior mean of $\phi$ is $\bar{\phi} =$ `r PhiBar` with 95% confidence interval (`r paste(lb, ub)`).

::: panel-tabset
## Exercise

Above we estimate $Var(\phi)$ to be `r round(PhiVar, 3)` and the standard error for $N = 100$ was `r round(se, 3)`.

If you wanted to state $p(\phi \in (\hat{\phi} \pm 0.01)) = 0.95$, how large would $N$ have to be? 

Check your answer by adjusting $N$ above.
:::

<!-- 2 sigma-hat / sqrt(N)  = 0.01 -->
<!-- ((sqrt(4.8) * 2) / 0.01)^2 -->