---
title: "MCMC Properties and Diagnostics"
author: "Dr. Alexander Fisher"
# mainfont: Lato
format: 
  html:
    toc: true
---

```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: View libraries used in these notes
library(tidyverse)
library(coda)
```

# Practice example

As in the example from lab [here](/slides/lab-normal.html#/model), let $y$ be the mercury concentration (ppm) of a bass fish and let $x$ be the weight of the fish (kg). This time, our model will be:

$$
\begin{aligned}
Y_i | \theta_1, \theta_2 &\sim N(\theta_1 + \theta_2 x_i, 1)\\
\theta_1, \theta_2 &\sim \text{ iid }N(0, 10)
\end{aligned}
$$

```{r}
#| label: load-data
#| warning: false
#| message: false
bass = read_csv("https://sta602-sp26.github.io/data/bass.csv")
y = bass$mercury
x = bass$weight
```

::: panel-tabset
## Exercise

Sample from the posterior $p(\theta_1, \theta_2 | y_1, \ldots, y_n, x_1, \ldots, x_n)$ using the Metropolis algorithm.

## Solution

```{r}
#| label: metropolis-example-1

logLikelihood = function(theta1, theta2) {
  mu = theta1 + (theta2 * x)
  sum(dnorm(y, mu, 1, log = TRUE))
}

logPrior = function(theta1, theta2) {
  dnorm(theta1, 0, sqrt(10), log = TRUE) + 
    dnorm(theta2, 0, sqrt(10), log = TRUE)
}

logPosterior = function(theta1, theta2) {
  logLikelihood(theta1, theta2) + logPrior(theta1, theta2)
}


THETA1 = NULL
THETA2 = NULL
accept1 = 0
accept2 = 0
S = 10000
theta1_s = 0
theta2_s = 0
for (s in 1:S) {
  
  ## propose and update theta1
  theta1_proposal = rnorm(1, mean = theta1_s, .5)
   log.r = logPosterior(theta1_proposal, theta2_s) - 
     logPosterior(theta1_s, theta2_s)
   
   if(log(runif(1)) < log.r)  {
    theta1_s = theta1_proposal
    accept1 = accept1 + 1 
   }
   
   THETA1 = c(THETA1, theta1_s)
   
   ## propose and update theta2
    theta2_proposal = rnorm(1, mean = theta2_s, .5)
   log.r = logPosterior(theta1_s, theta2_proposal) - 
     logPosterior(theta1_s, theta2_s)
   
   if(log(runif(1)) < log.r)  {
    theta2_s = theta2_proposal
    accept2 = accept2 + 1 
   }
   
   THETA2 = c(THETA2, theta2_s)
   
}
```

## Trace plots

```{r}
#| echo: false
df_theta = data.frame(theta1 = THETA1,
           theta2 = THETA2)

df_theta %>%
  ggplot(aes(x = 1:nrow(df_theta))) +
  geom_line(aes(y = THETA1, col = "theta1"), alpha = 0.8) +
  geom_line(aes(y = THETA2, col = "theta2"), alpha = 0.8) +
  labs(x = "iteration", y = "", color = "Parameter")
```

## Posterior summary

| Parameter  | Posterior Mean             | 95% CI                                                                                   |
|-----------------|-----------------|--------------------------------------|
| $\theta_1$ | `r round(mean(THETA1), 2)` | (`r round(quantile(THETA1, 0.025)[[1]],2)`, `r round(quantile(THETA1, 0.975)[[1]], 2)`)  |
| $\theta_2$ | `r round(mean(THETA2), 2)` | (`r round(quantile(THETA2, 0.025)[[1]], 2)`, `r round(quantile(THETA2, 0.975)[[1]], 2)`) |

## lm solution

Compare to `lm`:

```{r}
fit = lm(mercury ~ weight, data = bass)
summary(fit)
```
:::

# Why does it work?

## Ergodic theorem

Under what conditions does Metropolis-Hastings MCMC work?

**Ergodic theorem**: If $\{\theta^{(1)}, \theta^{(2)}, \ldots \}$ is an *irreducible*, *aperiodic* and *recurrent* Markov chain, then there is a unique probability distribution $\pi$ such that as $s \rightarrow \infty$,

-   $Pr(\theta^{(s)} \in \mathcal{A}) \rightarrow \pi(\mathcal{A})$ for any set $\mathcal{A}$;

-   $\frac{1}{S} \sum g(\theta^{(s)}) \rightarrow \int g(x) \pi(x) dx$.

## Definitions

### stationary distribution 

$\pi$ is called the **stationary distribution** of the Markov chain because if $\theta^{(s)} \sim \pi$ and $\theta^{(s+1)}$ is generated from the Markov chain starting at $\theta^{(s)}$, then $Pr(\theta^{(s+1)} \in \mathcal{A}) = \pi(\mathcal{A})$.

### irreducible

A chain is **reducible** if the state-space can be divided into non-overlapping sets (due to some $J$). In practice, the proposal $J(\theta^* | \theta^{(s)})$ needs to let us go from any value of $\theta$ to any other, eventually.

### aperiodic

We want our Markov chain to be **aperiodic**. A value $\theta$ is said to be **periodic** with period $k>1$ if it can only be visited every $k$th iteration. A Markov chain without periodic states is **aperiodic**.

### recurrent

A value $\theta$ is **recurrent** if we are guaranteed to return to it *eventually*.

# Diagnostics

## MCMC Vocabulary

-   **autocorrelation**: how correlated consecutive values in the Markov chain are. Mathematically, we compute the sample autocorrelation between elements in the sequence that are $t$ steps apart using

$$
\text{acf}_t(\boldsymbol{\phi}) = 
\frac{\frac{1}{S - t} \sum_{s = 1}^{S-t} (\phi_s - \bar{\phi})(\phi_{s+t} - \bar{\phi})}
{\frac{1}{S-1} \sum_{s = 1}^S (\phi_s - \bar{\phi})^2}
$$ where $\boldsymbol{\phi}$ is a sequence of length $S$ and $\bar{\phi}$ is the mean of the sequence. Practically, we use `acf` function in R. Example:

```{r}
acf(THETA1, plot = FALSE)
```

The higher the autocorrelation, the more samples we need to obtain a given level of precision for our approximation. One way to state how precise our approximation is, is with *effective sample size*.

-   **effective sample size** (ESS): intuitively this is the effective number of exact samples "contained" in the Markov chain (see [Betancourt 2018](https://arxiv.org/pdf/1701.02434.pdf)). For further reading on ESS, see [the stan manual](https://mc-stan.org/docs/reference-manual/effective-sample-size.html). In practice we use `coda::effectiveSize()` function to compute. Example:

```{r}
#| warning: false
coda::effectiveSize(THETA1)
```

More precisely, the **effective sample size** (ESS) is the value $S_{eff}$ such that

$$
Var_{MCMC}[\bar{\phi}] = \frac{Var[\phi]}{S_{eff}}.
$$ In words, it's the number of independent Monte Carlo samples necessary to give the same precision as the MCMC samples. For comparison, recall $Var_{MC}[\bar{\phi}] = Var[\phi]/S$

-   **Stationarity** is when samples taken in one part of the chain have a similar distribution to samples taken from other parts of the chain. Intuitively, we want the particle to move from our arbitrary starting point to regions of higher probability$^*$, then we will say it has *achieved stationarity*.

Traceplots are a great way to visually inspect whether a chain has **converged**, or *achieved stationarity*. In the traceplot from the previous lecture, we can see that samples from the beginning of the chain look very different than samples at the end for delta = 0.1.

$^*$ recall that probability is really a volume in high dimensions of parameter space, and so it is not enough for a pdf to evaluate to a high value, there must also be sufficient volume.

-   **Mixing**: how well the particle moves between sets of high probability. Some might refer to this as how well the particle sojourns across the "typical set" (regions of high probability).
