---
title: "The normal model"
author: "Dr. Alexander Fisher"
# mainfont: Lato
format: 
  html:
    toc: true
---

```{r}
#| warning: false
#| message: false
# load packages
library(tidyverse)
library(latex2exp)
```

## Background

### Definition and vocabulary


Let $Y$ be normally distributed with mean $\theta$ and variance $\sigma^2$. Mathematically,

$$
Y | \theta, \sigma^2  \sim N(\theta, \sigma^2).
$$
The density
$$
\begin{aligned}
p(y ~|~ \theta, \sigma^2 ) &= (2\pi\sigma^2)^{-1/2} e^{-\frac{1}{2\sigma^2} (y-\theta)^2},\\
y &\in \mathbb{R},\\
\theta &\in \mathbb{R},\\
\sigma &\in \mathbb{R}^+.
\end{aligned}
$$

#### location, scale

- $\theta$ is called the 'location' parameter
- $\sigma$ is called the 'scale' parameter

#### precision

Notice that every time $\sigma^2$ appears in the density, it is inverted. For this reason, we often consider the inverse variance $(\frac{1}{\sigma^2})$ as a quantity of interest. This quantity has a special name, **precision**. Intuitively, the precision tells us how close $y$ is to the mean $\theta$. (Large precision = small variance = closer).


### plots of normal densities

::: callout-warning
In `R`, the arguments of `pnorm`, `dnorm`, `rnorm` are the **mean** and **standard deviation** (not the variance!)
:::

::: panel-tabset

## plot

```{r}
#| warning: false
#| echo: false
set.seed(123)
N = 10000
y.mc = rnorm(N, mean = 3, sd = 2)

df = data.frame(y.mc)

df %>%
  ggplot(aes(x = y.mc)) +
  geom_histogram(aes(y = ..density..), alpha = 0.6, fill = 'steelblue') +
  stat_function(fun = dnorm, args = list(mean = 3, sd = 2), aes(color = "N(3, 4)")) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 1), aes(color = "N(0, 1)")) +
  theme_bw() +
  labs(x = "y", y = "density", title = "Normal densities",
       color = "")

```
## code

```{r}
#| eval: false
set.seed(123)
N = 10000
y.mc = rnorm(N, mean = 3, sd = 2)

df = data.frame(y.mc)

df %>%
  ggplot(aes(x = y.mc)) +
  geom_histogram(aes(y = ..density..), alpha = 0.6, fill = 'steelblue') +
  stat_function(fun = dnorm, args = list(mean = 3, sd = 2), aes(color = "N(3, 4)")) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 1), aes(color = "N(0, 1)")) +
  theme_bw() +
  labs(x = "y", y = "density", title = "Normal densities",
       color = "")
```

:::

## Bayesian inference

In general, we wish to make inference about $\theta$ and $\sigma^2$ after observing some data $y_1, \ldots y_n$ and thus are interested in the posterior $p(\theta, \sigma^2 | y_1, \ldots y_n)$. 
This is the standard task we have seen thus far, and requires us to specify a joint prior $p(\theta, \sigma^2)$. Below, we will work to find a class of conjugate priors over $\theta$ and $\sigma^2$.

We can break up the joint posterior into two pieces from the axioms of probability:

$$
p(\theta, \sigma^2 | y_1, \ldots y_n) = p(\theta | \sigma^2, y_1, \ldots, y_n)p(\sigma^2|y_1, \ldots y_n)
$$
This suggests that we calculate the joint posterior by:

- first finding the full conditional of $\theta$: $p(\theta| \sigma^2, \vec{y})$
- and then finding the marginal posterior of $\sigma^2$: $p(\sigma^2 | \vec{y})$,

where $\vec{y} = \{y_1, \ldots y_n\}$.

### The full conditional of $\theta$

By Bayes' theorem,

$$
p(\theta| \sigma^2, \vec{y}) \propto \underbrace{p(\vec{y} |\theta, \sigma^2)}_{\text{likelihood}} \underbrace{p(\theta|\sigma^2)}_{\text{prior}}.
$$
To arrive at the full conditional posterior of $\theta$, we must first specify a prior on $\theta$.

Considering we have a normal likelihood, what is a conjugate class of densities for $\theta$?

::: {.callout-tip collapse="true"}

## answer
$\theta | \sigma^2 \sim N(\mu_0, \tau_0^2)$ for some $\mu_0 \in \mathbb{R}$ and $\tau_0^2 \in \mathbb{R}^+$ is conjugate.
:::

With the conjugate prior, our full conditional posterior $\{ \theta| \sigma^2, \vec{y} \} \sim N(\mu_n, \tau_n^2)$ where 

$$
\begin{aligned}
\mu_n &= 
\frac{\frac{1}{\tau_0^2}\mu_0 + \frac{n}{\sigma^2} \bar{y}}{
\frac{1}{\tau_0^2} + \frac{n}{\sigma^2}
}
\\
\\
\tau_n^2 &= \frac{1}{\frac{1}{\tau_0^2} + \frac{n}{\sigma^2}}
\end{aligned}
$$

- Let's sketch out 'completing the square' to derive the parameters offline.

### Intuitive posterior parameters

If we consider the posterior precision, $\frac{1}{\tau_n^2}$, we can re-arrange the terms above to illuminate how posterior information = prior information + data information;.

$$
\frac{1}{\tau_n^2}= \frac{1}{\tau_0^2} + \frac{n}{\sigma^2}
$$
In words, posterior precision is equivalent to prior precision plus sampling precision. If we name each precision term, $\lambda_0 = \frac{1}{\tau_0}$ and $\lambda = \frac{1}{\sigma}$ then

$$
\mu_n = \frac{\lambda_0^2}{\lambda_0^2 + n\lambda^2} \mu_0 + 
\frac{n\lambda^2}{\lambda_0^2 + n\lambda^2} \bar{y}
$$

i.e. the posterior mean is the weighted average of prior and sample mean, where the weights are the relative contribution of each precision!

We can re-define $\lambda_0^2 = \kappa_0 \lambda^2$ (or equivalently $\tau_0^2 = \frac{\sigma^2}{\kappa_0}$) and obtain

$$
\begin{aligned}
\mu_n &= \frac{\kappa_0}{\kappa_0 + n} \mu_0 + \frac{n}{\kappa_0 + n} \bar{y},\\
\frac{1}{\tau_n^2} &= \frac{\kappa_0 + n}{\sigma^2}
\end{aligned}
$$

where we can interpret $\kappa_0$ as the prior sample size.

<!-- We can quickly write down the two terms of interest by remembering that any conditional, e.g. $p(\theta | \sigma^2, \vec{y})$ and $p(\sigma^2 | \theta, \vec{y})$ is proportional to the posterior (recall exercise 2.3 from Hoff on homework 1). -->

<!-- By Bayes' rule, the joint posterior -->

<!-- $$ -->
<!-- p(\theta, \sigma^2 | y_1, \ldots y_n) = \frac{p(y_1,\ldots y_n | \theta, \sigma^2)p(\theta, \sigma^2)}{p(y_1,\ldots y_n)}, -->
<!-- $$ -->
<!-- so the two distributions we wish to find can be written -->

<!-- - $p(\theta| \sigma^2, \vec{y}) \propto p(\vec{y}| \theta, \sigma^2) p(\theta |\sigma^2)$ -->
<!-- <!-- - $p(\sigma^2 | \vec{y}) \propto p(\vec{y}| \theta, \sigma^2) p(\sigma^2 |\theta)$ -->

<!-- and for the variance, -->

:::callout-important
Note that $\tau_n^2$ is the posterior variance of the full conditional posterior of $\theta$. This is distinct from $\sigma_n^2$, defined below.
:::


### Prior on $\sigma^2$

Remember, we want $p(\theta, \sigma^2 | \vec{y}) = p(\theta | \sigma^2, y_1, \ldots, y_n)p(\sigma^2|y_1, \ldots y_n)$. We have the first component of the right hand side, what about the second component?

Notice that 

$$
p(\sigma^2 | \vec{y}) \propto p(\sigma^2)\int p(\vec{y} | \theta, \sigma^2) p(\theta|\sigma^2) d\theta
$$

But how do we choose $p(\sigma^2)$ to be conjugate? We can proceed in multiple ways: one is noting that the integral is really a convolution of normals, (thereby a sum of normals) and is therefore a normal density.

Upon inspection, we can see a suitable choice is $\frac{1}{\sigma^2} \sim \text{gamma}(a, b)$. 

### The inverse-gamma

A random variable $X \in (0, \infty)$ has an inverse-gamma(a,b) distribution if $\frac{1}{X}$ has a gamma(a,b) distribution.

If X has an inverse-gamma distribution, the density of X is

$$
p(x | a, b) = \frac{b^a}{\Gamma(a)} x^{-a-1}e^{-b/x} \ \text{for } \ x > 0
$$
and
$$
\begin{aligned}
EX &= \frac{b}{(a-1)} \text{ if } a \geq 1; \ \infty \text{ if } 0<a<1,\\
Var(X) &= \frac{b^2}{(a-1)^2(a-2)} \ \text{if } a \geq 2; \ \infty \text{ if } 0 < a < 2,\\
Mode(X) &= \frac{b}{a +1}.
\end{aligned}
$$

### The marginal posterior of $\sigma^2$

Taken all together, if we let our sampling model and prior distributions be such that
$$
\begin{aligned}
Y_i | \theta, \sigma^2 &\sim N(\theta, \sigma^2)\\
\theta | \sigma^2 & \sim N(\mu_0, \sigma^2/\kappa_0)\\
\frac{1}{\sigma^2} &\sim \text{gamma}(\frac{\nu_0}{2}, \frac{\nu_0}{2} \sigma_0^2)
\end{aligned}
$$
then the posterior

$$
\frac{1}{\sigma^2} | \vec{y} \sim \text{gamma}(\frac{\nu_n}{2}, \frac{\nu_n \sigma^2_n}{2}),
$$
where 
$$
\begin{aligned}
\nu_n &= \nu_0 + n,\\
\sigma^2_n &= \frac{1}{\nu_n} \left[ 
\nu_0 \sigma^2_0 +(n-1)s^2 + \frac{\kappa_0 n}{\kappa_0 + n}(\bar{y} - \mu_0)^2
\right],
\end{aligned}
$$

and $s^2$ is the sample variance, $\frac{1}{n-1} \sum_i (y_i - \bar{y})^2$.


---

## Sampling from the joint posterior

Since $p(\theta, \sigma^2 | \vec{y}) = p(\theta | \sigma^2, y_1, \ldots, y_n)p(\sigma^2|y_1, \ldots y_n)$, we can sample from the joint posterior by first sampling from $p(\sigma^2|y_1, \ldots y_n)$ and then sampling from $p(\theta | \sigma^2, y_1, \ldots, y_n)$.

### Example

**Proof of concept**

We have some data:

```{r}
# generating 10 samples from the population
set.seed(123)
true.theta = 4
true.sigma = 1
y = rnorm(10, true.theta, true.sigma)

ybar = mean(y) # sample mean
n = length(y) # sample size
s2 = var(y) # sample variance
```

We make inference about $\theta$ and $\sigma^2$:

```{r}
# priors
# theta prior
mu_0 = 2; k_0 = 1
# sigma2 prior
nu_0 = 1; s2_0 = 0.010

# posterior parameters
kn = k_0 + n
nun = nu_0 + n
mun = (k_0 * mu_0 + n * ybar) /kn
s2n = (nu_0 * s2_0 + (n - 1) * s2 + k_0 * n * (ybar - mu_0)^2 / (kn)) / (nun)

s2.postsample = 1 / rgamma(10000, nun / 2, s2n * nun / 2)
theta.postsample = rnorm(10000, mun, sqrt(s2.postsample / kn))

df = data.frame(theta.postsample, s2.postsample)

df %>%
  ggplot(aes(x = theta.postsample, y = s2.postsample)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon") +
  labs(x = TeX("$\\theta$"),
       y = TeX("$\\sigma^2$"),
       fill = TeX("$p(\\theta, \\sigma^2 | y_1, \\ldots y_n)$")) +
  theme_bw()

```
