---
title: "Hierarchical modeling"
author: "Dr. Alexander Fisher"
# mainfont: Lato
format: 
  html:
    toc: true
---

```{r}
#| warning: false
#| code-summary: view packages used in these notes
#| code-fold: true
# load packages
library(tidyverse)
library(coda)
```

## Data set: school test scores

- Example from Hoff Ch. 8

Each year, students across North Carolina take an identical standardized test. In our sample, we observe scores from students at $m$ different schools. At the $j$th school, $n_j$ students take the exam and $j \in \{1, \ldots m\}$. The exam is designed to give an average score of 50 on a 0 to 100 scale.


```{r}
#| warning: false
# load data
mathScores = read_csv("https://sta602-sp26.github.io/data/mathScores.csv")
```

::: panel-tabset

## Codebook and glimpse

- `school`: which school the math score came from
- `mathscore`: score from 0 to 100 of an individual student

```{r}
#| echo: false
glimpse(mathScores)
```


## School scores

```{r}
#| warning: false
#| echo: false
mathScores %>%
  filter(school %in% c(1, 10, 41, 51)) %>%
  ggplot(aes(x = mathscore)) +
  geom_histogram() +
  theme_bw() +
  facet_wrap(vars(school)) +
  labs(title = "Distribution of raw math scores in four schools")
```

## All mean scores
```{r}
#| echo: false
#| warning: false
mathScores %>%
  group_by(school) %>%
  summarize(ybar = mean(mathscore)) %>%
  ggplot(aes(x = ybar)) +
  geom_histogram(bins = 12) +
  labs(x = "school-specific average score", y = "count", 
       title = "Distribution of school-specific average score") + 
  theme_bw()
```

:::

```{r}
#| code-fold: true
#| code-summary: "Convert data to list for downstream processing"
Y.school.mathscore <- as.matrix(mathScores)
#### Put data into list form.
Y <- list()
J <- max(Y.school.mathscore[, 1])
n <- ybar <- ymed <- s2 <- rep(0, J)
for (j in 1:J) {
  Y[[j]] <- Y.school.mathscore[Y.school.mathscore[, 1] == j, 2]
}
```

## Questions about the data

- How are the schools ranked?
- Does school 51 have a higher average score than school 41?
- What is the probability a single student randomly selected from school 51 performs better on the exam than a single student randomly selected from school 41?

## Model

Suppose students scores at school $j$ are *exchangeable* for all $n_j$. By de Finetti's theorem, this means 

$$
\{y_{1,j}, \ldots y_{n_j,j} | \phi_j \} \sim \text{ i.i.d. } p(y|\phi_j).
$$
That is, the student's scores at school $j$ are conditionally i.i.d. given some school specific parameters $\phi_j$. This describes our *within-group* sampling variability.

Now suppose that all the schools we sampled are similar in some way. Maybe they belong to some larger population of schools across the country i.e. schools in North Carolina are somewhat distinct from schools in South Carolina. We might imagine that the school-specific parameters themselves are exchangeable for all $m$. By de Finetti's theorem, this means

$$
\{\phi_1, \ldots \phi_m\} \sim \text{ i.i.d. } p(\phi|\psi).
$$

In words, school-specific parameters are conditionally i.i.d. given some population specific parameters $\psi$. This describes our *between-group* sampling variability.

Finally, if our hierarchy stops there, then to complete model specification, we may describe our prior beliefs about $\psi$ according to some prior density $p(\psi)$.

::: panel-tabset

## Exercise

Imagine variability among scores is the same across all schools, but there does exist heterogeneity in the mean scores of the schools. Write down the mathematical form of a  model that describes this using the normal distribution. What are some priors you could pick on relevant parameters to make sure full conditionals are easy to compute for Gibbs sampling? What are the full conditionals?

<!--comment out begins here 001-->

## Solution

- sampling distributions:

$$
\begin{aligned}
y_j | \theta_j, \sigma^2 &\sim N(\theta_j, \sigma^2)\\
\theta_j | \mu, \tau^2 &\sim N(\mu, \tau^2)
\end{aligned}
$$

- priors distributions:

$$
\begin{aligned}
1/\sigma^2 &\sim \text{ gamma}(\nu_0/2, \nu_0 \sigma_0^2/2)\\
1/\tau^2 &\sim \text{ gamma}(\eta_0/2, \eta_0 \tau_0^2/2)\\
\mu &\sim N(\mu_0, \gamma_0^2)
\end{aligned}
$$
:::
---

To facilitate Gibbs sampling, notice

$$
p(\theta_1, \ldots \theta_m, \mu, \tau^2, \sigma^2 | \mathbf{y}_1, \ldots \mathbf{y}_m) \propto
p(\mu, \tau^2, \sigma^2) p(\theta_1, \ldots \theta_m | \mu, \tau^2, \sigma^2) \times p(\mathbf{y}_1, \ldots \mathbf{y}_m| \theta_1, \ldots \theta_m, \mu, \tau^2, \sigma^2)
$$

It follows that the full conditionals are:

$$
\begin{aligned}
p(\mu | \cdot) &\propto p(\mu) \prod_{j = 1}^m p(\theta_j| \mu, \tau^2)\\
p(\tau^2 | \cdot) &\propto p(\tau^2) \prod_{j = 1}^m p(\theta_j| \mu, \tau^2)\\
p(\sigma^2|\cdot) &\propto p(\sigma^2)\prod_{j =1}^m \prod_{i = 1}^{n_j} p(y_{i,j}|\theta_j, \sigma^2)\\
p(\theta_j | \cdot) &\propto p(\theta_j | \mu, \tau^2) \prod_{i = 1}^{n_j} p(y_{i,j}|\theta_j, \sigma^2)
\end{aligned}
$$

<!--comment out ends here 001-->

## Gibbs sampling
```{r}
#### MCMC approximation to posterior for the hierarchical normal model

## weakly informative priors
nu0 <- 1; s20 <- 100
eta0 <- 1; t20 <- 100
mu0 <- 50; g20 <- 25

## starting values
m <- length(Y)
n <- sv <- ybar <- rep(NA, m)
for (j in 1:m)
{
  ybar[j] <- mean(Y[[j]])
  sv[j] <- var(Y[[j]])
  n[j] <- length(Y[[j]])
}
theta <- ybar
sigma2 <- mean(sv)
mu <- mean(theta)
tau2 <- var(theta)

## setup MCMC
set.seed(1)
S <- 5000
THETA <- matrix(nrow = S, ncol = m)
MST <- matrix(nrow = S, ncol = 3)
predictiveY = NULL

## MCMC algorithm
for (s in 1:S)
{
  # sample new values of the thetas
  for (j in 1:m)
  {
    vtheta <- 1 / (n[j] / sigma2 + 1 / tau2)
    etheta <- vtheta * (ybar[j] * n[j] / sigma2 + mu / tau2)
    theta[j] <- rnorm(1, etheta, sqrt(vtheta))
  }
  
  #sample new value of sigma2
  nun <- nu0 + sum(n)
  ss <- nu0 * s20
  for (j in 1:m) {
    ss <- ss + sum((Y[[j]] - theta[j]) ^ 2)
  }
  sigma2 <- 1 / rgamma(1, nun / 2, ss / 2)
  
  #sample a new value of mu
  vmu <- 1 / (m / tau2 + 1 / g20)
  emu <- vmu * (m * mean(theta) / tau2 + mu0 / g20)
  mu <- rnorm(1, emu, sqrt(vmu))
  
  # sample a new value of tau2
  etam <- eta0 + m
  ss <- eta0 * t20 + sum((theta - mu) ^ 2)
  tau2 <- 1 / rgamma(1, etam / 2, ss / 2)
  
  #store results
  THETA[s, ] <- theta
  MST[s, ] <- c(mu, sigma2, tau2)
  
  # predictive sampling
  y51 = rnorm(1, mean = theta[51], sd = sqrt(sigma2))
  y41 = rnorm(1, mean = theta[41], sd = sqrt(sigma2))
  predictiveY = rbind(predictiveY, c(y51, y41))
  
}

mcmc1 <- list(THETA = THETA, MST = MST)
```

## MCMC diagnostics

### trace plots

::: panel-tabset

## plots
```{r}
#| echo: false
colnames(MST) = c("mu", "sigma", "tau")
MST2 = MST %>%
  as.data.frame() %>% 
  pivot_longer(cols = 1:3)

MST2 %>%
  ggplot(aes(x = seq(1, nrow(MST2)), y = value)) +
  geom_line() +
  theme_bw() +
  facet_wrap(~ name, scales = "free_y") +
  labs(x = "iteration",
       title = "Traceplot of 5000 Gibbs samples")
```

## code

```{r}
#| eval: false
colnames(MST) = c("mu", "sigma2", "tau2")
MST2 = MST %>%
  as.data.frame() %>% 
  pivot_longer(cols = 1:3)

MST2 %>%
  ggplot(aes(x = seq(1, nrow(MST2)), y = value)) +
  geom_line() +
  theme_bw() +
  facet_wrap(~ name, scales = "free_y") +
  labs(y = "mu",
       x = "iteration",
       title = "Traceplot of 5000 Gibbs samples")
```
:::

### effective sample size and autocorrelation

```{r}
effectiveSize(MST)
par(mfrow=c(1,3))
acf(MST[,1])
acf(MST[,2]) 
acf(MST[,3]) 
```

### posterior means and standard error

```{r}
# MC error of mu, sigma2, tau2
MCERR <- apply(MST,2,sd)/sqrt( effectiveSize(MST) )
apply(MST,2,mean)
MCERR
```

We can do the exact same for the thetas, but the output will be 100 lines, so I suppress output below.

```{r}
#| eval: false
# MC error of thetas
effectiveSize(THETA) -> esTHETA
TMCERR <- apply(THETA,2,sd)/sqrt( effectiveSize(THETA) )
TMCERR
```


## Answers

- How are the schools ranked? How does the ordering compare to just ranking the schools by the sample means?

```{r}
# Ordering E[theta | data] and comparing to ybar

posteriorMean = THETA %>%
  apply(2, mean)

orderedTable = mathScores %>%
  group_by(school) %>%
  summarize(ybar = mean(mathscore),
            n = n()) %>%
  cbind(posteriorMean) %>%
  arrange(posteriorMean) %>%
  relocate(school, n, ybar, posteriorMean) %>%
  mutate_if(is.numeric, round, digits = 2)

DT::datatable(
  orderedTable,
  fillContainer = FALSE, options = list(pageLength = 10)
)
```

How many of the schools are ranked in the same position in the posterior ordering as the sample mean ordering? 

::: panel-tabset

## output
```{r}
#| echo: false
postOrdering = posteriorMean %>%
  order()

ybarOrdering = mathScores %>%
  group_by(school) %>%
  summarize(ybar = mean(mathscore), 
            n = n()) %>%
  arrange(ybar) %>%
  pull(school)

sum(postOrdering == ybarOrdering)
```

## code

```{r}
#| eval: false
postOrdering = posteriorMean %>%
  order()

ybarOrdering = mathScores %>%
  group_by(school) %>%
  summarize(ybar = mean(mathscore), 
            n = n()) %>%
  arrange(ybar) %>%
  pull(school)

sum(postOrdering == ybarOrdering)
```
:::

- Does school 51 have a higher average score than school 41? Re-cast as a Bayesian question: what's $p(\theta_{51} > \theta_{41} | \text{data})$?

```{r}
mean(THETA[,51] > THETA[,41])
```

- What's the probability a student randomly selected from school 51 performs better than a student selected randomly from school 41? 

Before looking at the solution below, how would you answer this problem? 

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Solution"
mean(predictiveY[,1] > predictiveY[,2])

# output:
# [1] 0.685
```

